|

### 2022-03-24 replication_paper/IntroducingICBe_DouglassEtAl_2021_BetaDraft_bookdown.Rmd
paper edits

If we record every event between countries in all of diplomacy, military conflict, and international political economy, how much information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could, in principle, be unbounded. Still, the central conceit of political science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured. Massive effort is expended in social science recording these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] With improvements in natural language processing, more open-ended efforts have sought to capture entire unstructured streams of international events drawn from news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations (IR). We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999;@brecherCrisisEscalationWar2000;@wilkenfeldInterstateCrisesViolence2000;@jamesWhatWeKnow2019;@iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state-of-the-art event data collection efforts. As should come as no surprise to anyone who works with event data, we find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs that we care about is unrecognizable through the lens of current quantification efforts.

@@

Successful measurement requires the existence of an unobserved latent state among all possible states, $\omega \in \Omega$ and a systematic data generating process (DGP) that converts that state into observables that enter the historical record, $X=f(\omega)$. We organize our prior beliefs about possible states, $\Omega$, and the DGP, $f()$, along two overarching axes, summarized in detail by Table 1. <!-- {tbl:litreview}  -->The first axes (rows) represents the types of information we expect to find in IR and forms the basis for our proposed ontology. We employ a metaphor of international affairs as a chess game, with players (polities, rebel groups, IGOs, etc.), pieces (military platforms, civilians, domains), and behaviors (think, say, do). Precise sequencing is required to capture gambits (sequences of moves) and outcomes (victory, defeat, peace, etc.), while precise geo-coding is required to understand the chessboard (medium of conflict). The resulting codebook for events includes 117 different verbs and is available in full in the online material.^[See the Github Repository [ICBEventData](https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$).]

The second axis (columns) represents sources of evidence for our concept priors and the existing state of the art. The first column ('Literature') maps each type of information to recent studies documenting its importance to IR. The second column ('ICB Corpus') documents the presence of that information in the ICB Corpus as identified by an expert review and natural language processing (Appendix \ref{app:verbs}). The third set of columns compare the ontologies of event-level datasets beginning with our contribution ICBe, then Cameo dictionary lookup based systems (Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018]; Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013]. Finally, the fourth set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010].

@@

Each crisis was assigned to at least two expert coders, 3 to 5 novice coders, with a third tie-breaking expert coder assigned to sentences with high disagreement. Expert coders were graduate students or postgraduates who collaboratively developed the ontology and documentation for the codebook, while novice coders were undergraduate students who engaged in classroom workshops. Each coder was instructed to thoroughly read and familiarize themselves with the full text. Once a coder felt that they understood the crisis, the coder was presented with a novel graphic user interface (GUI) that steps the coder through the narrative sentence by sentence (Figure \ref{fig:coding_gui}).

@@

One form of internal validity for these codings is high agreement between coders when applying the ontology. We employ three ways to evaluate and maximize agreement: (1) multiple expert and novice coders per sentence (often including a tie-breaking expert coder where there were disagreements), (2) a theoretically motivated scheme for aggregating across disagreements, and (3) survey questions on self-reported confidence and reasons for low confidence. The unit of analysis for these comparisons is the individually applied tag; every answer to every question in the ontology is a separate tag applied to that sentence (e.g., an actor, a behavior, a date, etc.). There were 113 coders (including 5 expert coders) who recorded `r icb_long_clean %>% nrow() %>% scales::comma()` coder-tags across the entire corpus. Of those, there were `r codings_long_agreement %>% filter(value_normalized!='') %>% nrow() %>% scales::comma()` unique sentence-tags applied by at least one coder. We ask how much support those possible tags received across coders, and how we should adjudicate disagreements when a tag was not chosen by every coder.

@@

This data-generating process suggests a theoretically motivated algorithm for aggregating over disagreements (Algorithm 1). We filter tokens to only those with (1) at least one expert coder vote and (2) a majority of expert votes or a majority of novice votes. If neither majority exists, we accept the one token with the most votes. This leaves `r codings_long_agreement %>% filter(value_normalized!='' & keep==T) %>% nrow() %>% scales::comma()` (`r (codings_long_agreement %>% filter(value_normalized!='' & keep==T) %>% nrow() / codings_long_agreement %>% filter(value_normalized!='') %>% nrow()) %>% round(2) %>% scales::percent()`) accepted tokens. In practice, this screens noisy tags that no expert considered possible but leverages novice knowledge to tie-break between equally plausible tags chosen by experts. A final step handles possible disagreement over the number of events or participants in the events. An event is emitted for each unique set of actors along with their surviving tags.

\begin{algorithm}[H]
\SetAlgoLined
\KwData{Coder-Sentence-Tokens}
\KwResult{Sentence-Events}
\ForEach{sentence in sentences}{
  \ForEach{concept in concepts}{
    \ForEach{token in tokens}{
      \lIf{expert votes==0}{reject; \tcc*[f]{Require at least one expert vote}}
      \uElseIf{expert votes > expert voters/2}{accept; \tcc*[f]{And a majority of expert votes}}
      \uElseIf{novice votes > novice voters/2}{accept; \tcc*[f]{Or a majority of novice votes}}
    }
    \If{sum(accepted) == 0}{
      sort tags by votes;
      head 1;
      accept; \tcc*[f]{If still no tag, take one with most votes}
    }
  }
  \ForEach{ActorSet in unique(ActorSets)}{
    emit event;
  }
}
\caption{Aggregating Coder-Tokens into Sentence-Events}
\end{algorithm}

@@

To demonstrate validity at the level of entire episodes, we conduct two qualitative case studies; we choose the Cuban Missile Crisis for being very high stakes and widely recognized and the DRC Civil War for being particularly complicated, lengthy, and modern. In addition, we provide visualizations for every crisis with side-by-side comparisons across datasets on the companion website. Our goals are as follows: (1) establish necessary and secondary details of each crisis from secondary historical sources independent of the ICB corpus; (2) evaluate the precision and overall coherence of ICBe codings using a type of timeline visualization we call a 'crisis map'; and (3) assess the recall of key events by the ICBe codings in comparison to the existing state-of-the-art datasets. For an event dataset to have good performance, it must both have high recall of salient facts, catching the key details and not just the trivia, and good precision, recording only events that actually occurred without distracting duplicates, miscodings, or other apparent examples of measurement error.

@@

By comparison, crisis maps of Cameo (Phoenix) (Figure \ref{fig:426metromap_phoenix}), ICEWS (Figure \ref{fig:426metromap_icews}), and Terrier (Figure \ref{fig:426metromap_terrier}), are unintelligible despite deduplication and heavy filtering of irrelevant or uninteresting events. The automated dictionary coding of news articles produces too many irrelevant events between unimportant actors or describes important events too vaguely to be used directly, and instead may only be interpretable in terms of weighted counts over time [@terechshenkoHotCollarLatent2020]. ICB Dyadic Events (Figure \ref{fig:426metromap_icb}), MIDs Incidents (Figure \ref{fig:426metromap_mids_incidents}), and MIDs Episodes (Figure \ref{fig:426metromap_mids}) do a good job showing the start and stop of conflict and between which actors, but lose any detailed context.

@@

We release several open-source products along with supporting code and documentation to further advance the study of IR, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts). Second, we provide a coder-level disaggregation with multiple codings of each sentence  by experts and novices that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here (www.crisisevents.org).

@@

We thank the many undergraduate coders for their patience and dedication. Thanks to the Center for Peace and Security Studies and its membership for comments. Thank you to the many research assistants that contributed to this project: Helen Chung, Daman Heer, Syeda ShahBano Ijaz, Anthony Limon, Erin Ling, Ari Michelson, Prithviraj Pahwa, Gianna Pedro, Tobias Stodiek, Yiyi 'Effie' Sun, Erin Werner, Lisa Yen, and Ruixuan Zhang. This project was supported by a grant from the Office of Naval Research [N00014-19-1-2491] and benefited from the Charles Koch Foundation's support for the Center for Peace and Security Studies.

>>>>

If we record every event between countries in all of diplomacy, military conflict, and international political economy, how much information would this chronicle amount to, and how surprised would we be to see something new? In other words, what is the entropy of international relations? This record could, in principle, be unbounded. Still, the central conceit of political science is that there are structural regularities that limit what actors can do, their best options, and even which actors are likely to survive [@brecherInternationalStudiesTwentieth1999; @reiterShouldWeLeave2015]. If so, then these events can be systematically measured. Massive effort is expended in social science recording these regularities.^[See work on crises [@brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020], militarized disputes [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018; @maozDyadicMilitarizedInterstate2019], wars [@sarkeesResortWar181620072010; @reiterRevisedLookInterstate2016], organized violence [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018], political violence [@raleighIntroducingACLEDArmed2010], sanctions [@felbermayrGlobalSanctionsData2020], trade [@barariDemocracyTradePolicy], and international agreements [@kinneDefenseCooperationAgreement2020; @owsiakInternationalBorderAgreements2018; @vabulasCooperationAutonomyBuilding2021], dispute resolution [@vabulasCooperationAutonomyBuilding2021; @frederickIssueCorrelatesWar2017], and diplomacy [@moyerWhatAreDrivers2020; @sechserMilitarizedCompellentThreats2011].] With improvements in natural language processing, more open-ended efforts have sought to capture entire unstructured streams of international events drawn from news reports.^[See @liComprehensiveSurveySchemabased2021; @haltermanExtractingPoliticalEvents2020; @brandtPhoenixRealTimeEvent2018; @boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020; @grantOUEventData2017. On event-extraction from images and social-media see @zhangCASMDeepLearningApproach2019 and @steinert-threlkeldFutureEventData2019.] How close these efforts are to accurately measuring all or even most of what is essential in international relations is an open empirical question, one for which we provide new evidence here.

Our contribution is a high coverage ontology and event dataset for key historical episodes in 20th and 21st-century international relations (IR). We develop a large, flexible ontology of international events with the help of both human coders and natural language processing. We apply it sentence-by-sentence to an unusually high-quality corpus of historical narratives of international crises [@brecherInternationalStudiesTwentieth1999; @brecherCrisisEscalationWar2000; @wilkenfeldInterstateCrisesViolence2000; @jamesWhatWeKnow2019; @iakhnisCrisesWorldPolitics2019]. The result is a new lower bound estimate of how much actually happens between states during pivotal historical episodes. We then develop several methods for objectively gauging how well these event codings reconstruct the information contained in the original narrative. We conclude by benchmarking our event codings against several current state-of-the-art event data collection efforts. As should come as no surprise to anyone who works with event data, we find that existing systems produce sequences of events that do not contain enough information to reconstruct the underlying historical episode. The underlying fine-grained variation in international affairs that we care about is unrecognizable through the lens of current quantification efforts.

@@

Successful measurement requires the existence of an unobserved latent state among all possible states, $\omega \in \Omega$ and a systematic data generating process (DGP) that converts that state into observables that enter the historical record, $X=f(\omega)$. We organize our prior beliefs about possible states, $\Omega$, and the DGP, $f()$, along two overarching axes, summarized in detail by Table 1. <!-- {tbl:litreview}  -->The first axes (rows) represents the types of information we expect to find in IR and forms the basis for our proposed ontology. We employ a metaphor of international affairs as a chess game, with players (polities, rebel groups, IGOs, etc.), pieces (military platforms, civilians, domains), and behaviors (think, say, do). Precise sequencing is required to capture gambits (sequences of moves) and outcomes (victory, defeat, peace, etc.), while precise geo-coding is required to understand the chessboard (medium of conflict). The resulting codebook for events includes 117 different behaviors and is available in full in the online material.^[See the Github Repository [ICBEventData](https://urldefense.com/v3/__https://github.com/CenterForPeaceAndSecurityStudies/ICBEventData__;!!Mih3wA!WxDJtEczKfxGTh0S2Krunap8ReymFEL5iTWaSfOHeqlSdyfRx77zmjBSWO1OAm13$).]

The second axis (columns) represents sources of evidence for our concept priors and the existing state of the art. The first column ('Literature') maps each type of information to recent studies documenting its importance to IR. The second column ('ICB Corpus') documents the presence of that information in the ICB Corpus as identified by an expert review and natural language processing (Appendix \ref{app:verbs}). The third set of columns compare the ontologies of event-level datasets beginning with our contribution ICBe, then Cameo dictionary lookup based systems (Real-Time Phoenix [@brandtPhoenixRealTimeEvent2018]; Historical Phoenix [@althausClineCenterHistorical2019]; ICEWS [@boscheeICEWSCodedEvent2015; @hegreIntroducingUCDPCandidate2020]; Terrier [@grantOUEventData2017]), the Militarized Interstate Disputes Incidents dataset, and the UCDP-GED dataset [@ralphsundbergUCDPGEDCodebook2016; @petterssonOrganizedViolence19892018; @sundbergIntroducingUCDPGeoreferenced2013].^[Absent datasets include BCOW [@lengMilitarizedInterstateCrises1988], WEIS [@mcclellandWorldEventInteraction1978], CREON [@hermannComparativeResearchEvents1984], CASCON [@bloomfieldCASCONIIIComputeraided1989], and SHERFACS [@shermanSHERFACSCrossParadigmHierarchical2000] (see histories [@merrittMeasuringEventsInternational1994] and [@schrodtTwentyYearsKansas2006]).] Finally, the fourth set of columns compares episode-level datasets beginning with the original ICB project [@brecherInternationalCrisisBehavior; @brecherCrisesWorldPolitics1982; @beardsleyInternationalCrisisBehavior2020]; the Militarized Interstate Disputes dataset [@palmerMID5Dataset20112021; @giblerInternationalConflicts181620102018;@braithwaiteMIDLOCIntroducingMilitarized2010;@braithwaiteCodebookMilitarizedInterstate2009], and the Correlates of War [@sarkeesResortWar181620072010].

@@

Each crisis was assigned to at least two expert coders, 3 to 5 undergrad coders, with a third tie-breaking expert coder assigned to sentences with high disagreement.^[Expert coders were graduate students or postgraduates who collaboratively developed the ontology and documentation for the codebook. Undergrad coders were students who engaged in classroom workshops. Coders worked independently with the codebook after a practice coding where an expert answered questions.] Each coder was instructed to thoroughly read and familiarize themselves with the full text. Once a coder felt that they understood the crisis, the coder was presented with a novel graphic user interface (GUI) that steps the coder through the narrative sentence by sentence (Figure \ref{fig:coding_gui}).

@@

Without an existing validation dataset, we cannot use traditional measures of validity. We instead examine agreement between coders, juxtaposition with source sentences, and case study comparisons with peer datasets. One form of internal validity for these codings is high agreement between coders when applying the ontology. We employ three ways to evaluate and maximize agreement: (1) multiple expert and undergrad coders per sentence (often including a tie-breaking expert coder where there were disagreements), (2) a theoretically motivated scheme for aggregating across disagreements, and (3) survey questions on self-reported confidence and reasons for low confidence. The unit of analysis for these comparisons is the individually applied tag; every answer to every question in the ontology is a separate tag applied to that sentence (e.g., an actor, a behavior, a date, etc.). There were 113 coders (including 5 expert coders) who recorded `r icb_long_clean %>% nrow() %>% scales::comma()` coder-tags across the entire corpus. Of those, there were `r codings_long_agreement %>% filter(value_normalized!='') %>% nrow() %>% scales::comma()` unique sentence-tags applied by at least one coder. We ask how much support those possible tags received across coders, and how we should adjudicate disagreements when a tag was not chosen by every coder.

@@

This data-generating process suggests a theoretically motivated algorithm for aggregating over disagreements (Algorithm 1). We filter tokens to only those with (1) at least one expert coder vote and (2) a majority of expert votes or a majority of undergrad votes. If neither majority exists, we accept the one token with the most votes. This leaves `r codings_long_agreement %>% filter(value_normalized!='' & keep==T) %>% nrow() %>% scales::comma()` (`r (codings_long_agreement %>% filter(value_normalized!='' & keep==T) %>% nrow() / codings_long_agreement %>% filter(value_normalized!='') %>% nrow()) %>% round(2) %>% scales::percent()`) accepted tokens. In practice, this screens noisy tags that no expert considered possible but leverages undergrad knowledge to tie-break between equally plausible tags chosen by experts. A final step handles possible disagreement over the number of events or participants in the events. An event is emitted for each unique set of actors along with their surviving tags.

\begin{algorithm}[H]
\SetAlgoLined
\KwData{Coder-Sentence-Tokens}
\KwResult{Sentence-Events}
\ForEach{sentence in sentences}{
  \ForEach{concept in concepts}{
    \ForEach{token in tokens}{
      \lIf{expert votes==0}{reject; \tcc*[f]{Require at least one expert vote}}
      \uElseIf{expert votes > expert voters/2}{accept; \tcc*[f]{And a majority of expert votes}}
      \uElseIf{undergrad votes > undergrad voters/2}{accept; \tcc*[f]{Or a majority of undergrad votes}}
    }
    \If{sum(accepted) == 0}{
      sort tags by votes;
      head 1;
      accept; \tcc*[f]{If still no tag, take one with most votes}
    }
  }
  \ForEach{ActorSet in unique(ActorSets)}{
    emit event;
  }
}
\caption{Aggregating Coder-Tokens into Sentence-Events}
\end{algorithm}

@@

To demonstrate validity at the level of entire episodes, we conduct two qualitative case studies. We choose cases where Phoenix had substantially more relevant events (included crisis actors in crisis time span) then ICBe, likely exacerbating the well-known poor coherence of automated codings but increasing the likelihood of capturing key events. The two cases also differ in useful ways; the Cuban Missile Crisis is high stakes and widely recognized and the DRC Civil War for being particularly complicated, lengthy, and modern.

For each case we first establish necessary and secondary details of each crisis from secondary historical sources independent of the ICB corpus. We then evaluate the precision and overall coherence of ICBe codings using a type of timeline visualization we call a 'crisis map'.^[The project website has such visualizations for every crisis.]. Finally we compare the recall of key events by ICBe and the existing state-of-the-art datasets. For an event dataset to have good performance, it must both have high recall of salient facts, catching the key details and not just the trivia, and good precision, recording only events that actually occurred without distracting duplicates, miscodings, or other apparent examples of measurement error.

@@

By comparison, crisis maps of Cameo (Phoenix) (Figure \ref{fig:426metromap_phoenix}), ICEWS (Figure \ref{fig:426metromap_icews}), and Terrier (Figure \ref{fig:426metromap_terrier}), are unintelligible despite deduplication and heavy filtering of irrelevant or uninteresting events. The real-time of advantage of automated dictionary coding of news articles comes at the price of producing too many irrelevant events between unimportant actors or describes important events too vaguely to be used directly, and instead may only be interpretable in terms of weighted counts over time [@terechshenkoHotCollarLatent2020]. ICB Dyadic Events (Figure \ref{fig:426metromap_icb}), MIDs Incidents (Figure \ref{fig:426metromap_mids_incidents}), and MIDs Episodes (Figure \ref{fig:426metromap_mids}) do a good job showing the start and stop of conflict and between which actors, but lose any detailed context.

@@

We release several open-source products along with supporting code and documentation to further advance the study of IR, event extraction, and natural language processing. The first is the International Crisis Behavior Events (ICBe) dataset, an event-level aggregation of what took place during the crises identified by the ICB project. These data are appropriate for statistical analysis of hard questions about the sequencing of events (e.g. escalation and de-escalation of conflicts). Second, we provide a coder-level disaggregation with multiple codings of each sentence by experts and undergrads that allows for the introduction of uncertainty and human interpretation of events. Further, we release a direct mapping from the codings to the source text at the sentence level as a new resource for natural language processing. Finally, we provide a companion website that incorporates detailed visualizations of all of the data introduced here (www.crisisevents.org).

@@

We thank the many undergraduate coders for their patience and dedication. Thanks to the Center for Peace and Security Studies and its membership for comments. Special thanks to Rebecca Cordell, Philip Schrodt, Zachary Steinert-Threlkeld, and Zhanna Terechshenko for generous feedback. Thank you to the cPASS research assistants that contributed to this project: Helen Chung, Daman Heer, Syeda ShahBano Ijaz, Anthony Limon, Erin Ling, Ari Michelson, Prithviraj Pahwa, Gianna Pedro, Tobias Stodiek, Yiyi 'Effie' Sun, Erin Werner, Lisa Yen, and Ruixuan Zhang. This project was supported by a grant from the Office of Naval Research [N00014-19-1-2491] and benefited from the Charles Koch Foundation's support for the Center for Peace and Security Studies.
